---
title: "sea birds"
author: "Atalanta Ritter"
date: "2023-02-16"
output: pdf_document
---
## Sea Bird Data Analysis 
The chunk below is just for set up/loading in packages. 
```{r}
# set working directory
rm(list = ls())
getwd()
setwd("~/GitHub/team3")
# load in packages
package.list <- c('vegan', 'tidyverse', 'ggplot2', 'dplyr', 'broom', 'viridis')
for (package in package.list) {
  if (!require(package, character.only = TRUE, quietly = TRUE)) {
    install.packages(package)
  }
  library(c(package), character.only = TRUE)
}
```

## Cluster Analysis  
```{r}
# loading in dataset
seabirds <- read.csv("/Users/Atalanta/Downloads/Data_DRYAD_Seabirds.csv")
dataspecies <- seabirds[-c(37,59, 75,76,78,79,80, 82,83,112,121,135,165,172,177,190,198,259,280,307,311,312) ,2:5] 
# removed xtra columns, sites with no observations of any species and 
# one outlier site with 160 shearwaters
# cluster analysis
gplots::heatmap.2(as.matrix(dataspecies), 
                  distfun = function(x) vegdist(x, method = "bray"),
                  hclustfun = function(x) hclust(x, method = "ward.D2"),
                  col = viridis, trace = "none", density.info = "none")
```
After removing that one site with 160 shearwaters, it looks like we have some clearer clustering patterns! 

## Mantel Tests 
The code below is for a Mantel Test comparing the species data with data on distance from rat-infested vs. rat-free islands. 
```{r}
# creating matrix of distances from rat-infested and rat-free islands
datarat <- seabirds[-c(37,59, 75,76,78,79,80, 82,83,112,121,135,165,172,177,190,198,259,280,307,311,312) , c(20,22)] 
# mantel test: test whether there is a correlation between 
seabird.dist <- vegdist(dataspecies, method = "bray")
rat.dist <- vegdist(datarat, method = "euclid")
mantel(seabird.dist, rat.dist)
```
The results are significant (r = 0.1263, p = 0.001), meaning that as species composition becomes more dissimilar between sites, distances from rat-free and rat-infested islands also become more dissimilar. 

The code below is for two Mantel tests comparing species data with standardized environmental data on 4 different variables: sea surface temperature, chlorophyll a concentration, net primary productivity, and sea-level anomaly. 
We subdivided these environmental variables into two groups: food availability (NPP, CHL) and climate (SST, SLA). 
```{r}
# MANTEL TEST COMPARING PROXIES FOR FOOD AVAILABILITY VS. SPECIES DISTRIBUTION
datafood <- seabirds[-c(37,59, 75,76,78,79,80, 82,83,112,121,135,165,172,177,190,198,259,280,307,311,312) ,c(14,16)] 
# standardizing environmental data: scaled environmental variables to have 
# mean = 0 and standard deviation = 1 
datafood.std <- dataenv %>% mutate_at(c('Chl_Climonth', 'NPP_Climonth'), ~(scale(.) %>% as.vector))
# run mantel test 
food.dist <- vegdist(datafood.std, method = "euclid")
mantel(seabird.dist, food.dist)
# marginally significant, p = 0.065
# MANTEL TEST COMPARING CLIMATE VS. SPECIES DISTRIBUTION
dataclim <- seabirds[-c(37,59, 75,76,78,79,80, 82,83,112,121,135,165,172,177,190,198,259,280,307,311,312) ,c(15,17)] 
# standardizing environmental data: scaled environmental variables to have 
# mean = 0 and standard deviation = 1 
dataclim.std <- dataenv %>% mutate_at(c('SST_Climonth', 'SLA_Climonth'), ~(scale(.) %>% as.vector))
# run mantel test 
clim.dist <- vegdist(dataclim.std, method = "euclid")
mantel(seabird.dist, clim.dist)
# marginally significant, p = 0.069
```
## Linear Models
The Mantel test results suggest that distance from rat-infested islands is related to species distribution. Next, we wanted to ask the question of whether this trend is the same across seabird species or if the effects are species-specific. We ran linear models for each of the species vs. distance to nearest rat-infested island.

```{r}
# are the correlations between distance from rat islands + species distribution general or species specific? 
# linear regressions for rat distance vs. [species name] distribution
# 1. Brown Noddy
noddy.reg <- lm(Brown_Noddy ~ Dist_Coast_R, data = seabirds)
summary(noddy.reg)
# Brown Noddy plot 
ggplot(seabirds,aes(Dist_Coast_R, Brown_Noddy)) +
  geom_point() +
  geom_smooth(method='lm') +
  xlab("Distance to closest rat-infested island (km)") +
  ylab("Brown noddy abundance")
# highly significant (but negative??), p = 5.19e-8
# 2. Wedge-tailed Shearwater regression + plot 
shearwater.reg <- lm(Wedge_Tailed_Shearwater ~ Dist_Coast_R, data = seabirds[-80,])
summary(shearwater.reg) # get rid of 1 outlier site (doesn't change the significance that much)
ggplot(seabirds[-80,],aes(Dist_Coast_R, Wedge_Tailed_Shearwater)) +
  geom_point() +
  geom_smooth(method='lm') +
  xlab("Distance to closest rat-infested island (km)") +
  ylab("Wedge-Tailed Shearwater Abundance")
# no significance, p = 0.692
# 3. White Tern regression + plot 
tern.reg <- lm(White_Tern ~ Dist_Coast_R, data = seabirds[-c(99,238),]) # getting rid of outlier sites 
summary(tern.reg)
ggplot(seabirds[-c(99,238),],aes(Dist_Coast_R, White_Tern)) +
  geom_point() +
  geom_smooth(method='lm') +
  xlab("Distance to closest rat-infested island (km)") +
  ylab("White tern abundance")
# significant (again negative), p = 1.04e-5
# Red-footed Booby regression + plot 
booby.reg <- lm(Red_footed_Booby ~ Dist_Coast_R, data = seabirds[-144,])
summary(booby.reg)
ggplot(seabirds[-144,], aes(Dist_Coast_R, Red_footed_Booby)) +
  geom_point() +
  geom_smooth(method='lm') +
  xlab("Distance to closest rat-infested island (km)") +
  ylab("Red-footed booby abundance")
# not significant, p =0.305
```

