---
title: "sea birds"
author: "Atalanta Ritter"
date: "2023-02-16"
output: pdf_document
---
## Sea Bird Data Analysis 
The chunk below is just for set up/loading in packages. 
```{r}
# set working directory
rm(list = ls())
getwd()
setwd("~/GitHub/QB2023_Ritter")
# load in packages
package.list <- c('vegan', 'tidyverse', 'ggplot2', 'dplyr', 'broom')
for (package in package.list) {
  if (!require(package, character.only = TRUE, quietly = TRUE)) {
    install.packages(package)
  }
  library(c(package), character.only = TRUE)
}
```

## Cluster Analysis  
```{r}
# loading in dataset
seabirds <- read.csv("/Users/Atalanta/Downloads/Data_DRYAD_Seabirds.csv")
dataspecies <- seabirds[-c(37,59, 75,76,78,79,80, 82,83,112,121,135,165,172,177,190,198,259,280,307,311,312) ,2:5] 
# removed xtra columns, sites with no observations of any species and 
# one outlier site with 160 shearwaters
# cluster analysis
gplots::heatmap.2(as.matrix(dataspecies), 
                  distfun = function(x) vegdist(x, method = "bray"),
                  hclustfun = function(x) hclust(x, method = "ward.D2"),
                  col = viridis, trace = "none", density.info = "none")
```
After removing that one site with 160 shearwaters, it looks like we have some clearer clustering patterns! 

## Mantel Test 
The code below is for a Mantel Test comparing the species data with data on distance from rat-infested vs. rat-free islands. 
```{r}
# creating matrix of distances from rat-infested and rat-free islands
datarat <- seabirds[-c(37,59, 75,76,78,79,80, 82,83,112,121,135,165,172,177,190,198,259,280,307,311,312) , c(20,22)] 
# mantel test: test whether there is a correlation between 
seabird.dist <- vegdist(dataspecies, method = "bray")
rat.dist <- vegdist(datarat, method = "euclid")
mantel(seabird.dist, rat.dist)
```
The results are significant (r = 0.1263, p = 0.001), meaning that as species composition becomes more dissimilar between sites, distances from rat-free and rat-infested islands also become more dissimilar? 

## Mantel Test 
The code below is for a Mantel test comparing species data with standardized environmental data on 4 different variables: sea surface temperature, chlorophyll a concentration, net primary productivity, and sea-level anomaly. 
```{r}
# creating matrix with just the environmental data 
dataenv <- seabirds[-c(37,59, 75,76,78,79,80, 82,83,112,121,135,165,172,177,190,198,259,280,307,311,312) ,14:17] 
# standardizing environmental data: scaled environmental variables to have 
# mean = 0 and standard deviation = 1 
dataenv.std <- dataenv %>% mutate_at(c('Chl_Climonth', 'SST_Climonth', 'NPP_Climonth', 'SLA_Climonth'), ~(scale(.) %>% as.vector))
# run mantel test 
env.dist <- vegdist(dataenv.std, method = "euclid")
mantel(seabird.dist, env.dist)
```
This result is NOT significant (r = 0.02354, r = 0.12). 
